---
title: "Preprossessing data analysis for Iron and gut microbiome project data"
output: html_document
date: "2026-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # send any figure output to this folder 
                      fig.path = "../figures/02_PreProcessing/")
# Set the seed for reproducibility
set.seed(238428)

# What time did we start running this script? 
start_time <- Sys.time()

# Load packages 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, patchwork, install = FALSE)

# Load the data 
load("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/02_AssignASVs/raw_physeq.RData")

# Look at the data 
raw_physeq

```

```{r Clean up the data}
# Let's see where chloroplasts are and also mitochondria

# How to access what is in a phyloseq table
view(tax_table(raw_physeq))
view(otu_table(raw_physeq))

# Make new physeq without chloroplasts
noChloros_physeq <- 
  raw_physeq %>% 
  subset_taxa(Order != "Chloroplast" | is.na(Order))
  
# How many taxa were chloroplasts? 
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs

# remove mitochondria 
noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASVs? 
num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs

# Let's take a quick look 
noChlorosMitos_physeq

# How many total asvs were removed from chloros and mitos? 
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)

# proportion of asvs kept? 
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```

### Interpretation 1: 
In this project dataset, there are no chlorophyll and mitochondria ASV being detected. This is not superising, as they used a different primer set for V5-V6 which is less commonly applied for bacterial 16s sequencing. I think more generally/prevalently used primers might give higher detection rate of those contaminants, and the results seeing here is suggesting this dataset has good primer specificity with no host/dietary carryover.

### Interpretation 2 and 3:
This project dataset has no negative control and no mock community. It's harder to assess the accuracy of the sequencing, this is certainly a limitation in their microbial sequencing study design.


```{r Reverse Complements}
# Pull out Taxa Table
tax_table <- 
  noChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 
```
### Interpretation 4: 
This dataset do no have any reverse compliments to be removed. 

```{r}
# The current data object
noChlorosMitos_physeq

# Make a metadata_df for joining
metadata_df <- 
  noChlorosMitos_physeq %>%
  sample_data() %>% 
  data.frame()

library(tibble)

metadata_df <- metadata_df %>%
  rownames_to_column(var = "names")

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  noChlorosMitos_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_df, by = "names") 

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)

head(seqSums_df)
```

```{r}
# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()

numSeq_bar_plot <- 
  seqSums_df %>%
  ggplot(aes(x = reorder(names, TotalSeqs), y = TotalSeqs, fill = Group)) + 
  geom_col() + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample") + 
  theme(legend.position = "none")

numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x = TotalSeqs, fill = Group)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme(legend.position = "bottom")

numSeq_density_plot <- 
  seqSums_df %>%
  ggplot(aes(x = TotalSeqs, fill = Group)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth") + 
  theme(legend.position = "none")

numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A")
```
```{r Remove sample with few reads}
# What's the min seq depth? 
min(sample_sums(noChlorosMitos_physeq))

# Final check of the sequencing depth for now
# Here, we could remove more samples if we needed to
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)

preprocessed_physeq <- noChlorosMitos_physeq
```
### Interpretation 5: 
The minimum reads is 21454 which shall be good for 16s. The greatest read is in 60K which is ~3-fold greater than the minimal so rarefaction is recommended for downstream comparison. While the read depths for all samples are generally good, the sequencing depth slightly differed between treatment groups with the CRC groups getting more reads than the HC (healthy controls), which shall be considered when comparing diversity. But rarefactino will help with this when comparing differential abundances. The overall sequencing depth is good in general. 

```{r add a metadata column}
# Let's add a metadata column to the file for this dataset
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame() %>%
  rownames_to_column(var = "SampleID") %>%
  mutate(
    Iron_dose_mg = as.numeric(Iron.intake),
    Group = factor(Group,
                   levels = c("HC", "CRC"),
                   ordered = TRUE),
    Iron.intake = factor(Iron.intake,
                         levels = c(50, 500),
                         labels = c("Low iron", "High iron"),
                         ordered = TRUE),
    Group_Iron = factor(
      paste(Group, Iron.intake, sep = "_"),
      levels = c("HC_Low iron", "HC_High iron",
                 "CRC_Low iron", "CRC_High iron"),
      ordered = TRUE
    ),
    ID = factor(ID, ordered = TRUE)
  )

### Let's add the sequencing count data, so that doesn't get lost! 
# What are the sample size distributions?# Per-sample sequencing depth
num_seqs_df <- 
  preprocessed_physeq %>%
  sample_sums() %>%
  data.frame() %>%
  rownames_to_column(var = "SampleID")

colnames(num_seqs_df)[2] <- "Sequencing_Depth"

# Merge sequencing depth into metadata
metadata_df2 <- 
  metadata_df %>%
  left_join(num_seqs_df, by = "SampleID")

## Check they match
stopifnot(metadata_df$SampleID == metadata_df2$SampleID)

# Rename the rows 
rownames(metadata_df2) <- metadata_df2$SampleID

# Reassign to the phyloseq object, rewriting the old metadata 
sample_datapreprocessed_physeq <- metadata_df2

# Double check
colnames(data.frame(sample_data(preprocessed_physeq)))

# Intuition check 
head(num_seqs_df)
```
```{r save data}
# Please be sure to use the correct path to the appropriate directory in your file system
save(preprocessed_physeq, file = "/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/03_Pre-processing/preprocessed_phyloseq.RData")
```

```{r Ending}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time

# Ensure reproducibility with package version information
devtools::session_info()
```
