---
title: "Beta-Diversity"
output: html_document
date: "2026-04-15"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inputs:
- preprocessed_physeq: asv, count table, metadata, taxonomy

# Outputs
- visualize data in ordinations (PCoA or NMDs)
-Calculate stats (PERMANOVA and betadisper)

```{r Load data and packages}
pacman::p_load(tidyverse, phyloseq, patchwork, vegan,
               install = FALSE)

# Load colors and shapes
source("/local/workdir/yz3272/Biodiversity_Salinity_16S/code/colors_and_shapes.R")
source("/local/workdir/yz3272/Biodiversity_Salinity_16S/code/functions.R")

#load data
load("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/03_Pre-processing/preprocessed_phyloseq.RData")
preprocessed_physeq

# Intuition check on seq depth
min(sample_sums(preprocessed_physeq))
```


```{r Create metadata_df}
metadata_df <- 
  preprocessed_physeq %>%
  sample_data() %>%
  data.frame()
```

```{r Explore the Raw Read Counts}
# calculate read counts per sample 
raw_TotalSeqs_df <- 
  preprocessed_physeq %>%
  # Calculate the total number of sequences/reads
  sample_sums() %>%
  data.frame()

# Take a look 
head(raw_TotalSeqs_df)

# Rename the column 
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
raw_TotalSeqsASVs_df <- 
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(preprocessed_physeq) > 1))

#View(raw_TotalSeqsASVs_df)

# Plot histogram of seq depth 
rawSeq_histogram <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  scale_x_continuous(limits = c(0, 70000)) + 
  labs(title = "Raw Seq Depth Histogram") + 
  theme_bw()

# Plot Seq Depth versus num_ASVs
rawSeq_vs_numASV_plot <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 70000)) +
  geom_smooth(method = "lm") + 
  labs(title = "Seq Depth vs # ASVs") + 
  theme_bw()

# is it significant? 
summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))

# Put the two plots together.
rawSeq_histogram + rawSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```
```{r  Scale Read Counts}
min(sample_sums(preprocessed_physeq))

# Scale the reads 
scaled_physeq <- 
  preprocessed_physeq %>%
  scale_reads(round = "matround")

# Look at it 
scaled_physeq

# Look at it more!
#View(data.frame(otu_table(preprocessed_physeq)))
#View(data.frame(otu_table(scaled_physeq)))

# Confirm seq read depth of scaled_physeq 
scaled_TotalSeqs_df <- 
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
scaled_TotalSeqsASVs_df <- 
  scaled_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(scaled_physeq) > 1))

# Plot it! 
scaledSeq_histogram <- 
  scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  labs(title = "Scaled Seq Depth Histogram") + 
  scale_x_continuous(limits = c(0, 70000)) + 
  theme_bw()
scaledSeq_histogram

# Scaling always gives us a range! 
# MInimum sequences
min(scaled_TotalSeqs_df$TotalSeqs) 

# Total Range 
range(scaled_TotalSeqs_df) # a range happen due to the scaling factor differ for each sample
# How many is that range? 
range_seqs <- (max(scaled_TotalSeqs_df$TotalSeqs) - min(scaled_TotalSeqs_df$TotalSeqs))
range_seqs
range_seqs/max(scaled_TotalSeqs_df$TotalSeqs)

#View(scaled_TotalSeqs_df)

# Set the axis min and max for the next plot 
max_y <- max(scaled_TotalSeqs_df$TotalSeqs) + 10
min_y <- min(scaled_TotalSeqs_df$TotalSeqs) - 10

# Now, draw the plot 
scaledSeq_vs_numASV_plot <- 
  scaled_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(min_y, max_y)) +
  #geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(title = "Scaled: Seq Depth vs # ASVs")

# Put the two plots together.
scaledSeq_histogram + scaledSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```
###INTERPRETATION 3: Evaluate the scaled read count plots above. Are the total number of sequences across all of your samples in the same general range? What is the minimum and maximum (i.e., the range) of the sequencing depth? How was the relationship with the number of ASVs modified after scaling the read counts?

```{r Calculating dissimilarity}
# Sorensen Dissimiliarty
scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

# What does it look like? 
class(scaled_sorensen_dist)

str(scaled_sorensen_dist)

#head(as.matrix(scaled_sorensen_dist))

# Bray-Curtis Dissimiliarty
scaled_bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)

# Abundance-Unweighted UniFrac
#scaled_uUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "unifrac")

# Abundance-Weighted UniFrac
#scaled_wUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "wunifrac")
```

### INTERPRETATION 4: Which dissimilarity metric is most related to your scientific question? Why or why not?

```{r Sorensen}
## 1. Run with by = terms for R² values, sensitive to order of variables! 
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
sorensen_station_adonis_terms1 <- adonis2(scaled_sorensen_dist ~ Group + Iron_dose_mg, data = metadata_df2, by = "terms")
sorensen_station_adonis_terms1

# Check the order of the terms
sorensen_station_adonis_terms2 <- adonis2(scaled_sorensen_dist ~ Group + Iron_dose_mg, data = metadata_df2, by = "terms")
sorensen_station_adonis_terms2

## 2. Run with by = "margin" for marginal p-values, which we can compare to the residuals from the first one. 
sorensen_station_adonis_margin <- adonis2(scaled_sorensen_dist ~ Group + Iron_dose_mg, data = metadata_df2, by = "margin")
sorensen_station_adonis_margin
```

```{r Bray-Curtis}
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
bray_station_adonis_terms1 <- adonis2(scaled_bray_dist ~ Iron_dose_mg, data = metadata_df2, by = "terms")
bray_station_adonis_terms1

bray_station_adonis_terms2 <- adonis2(scaled_bray_dist ~ Iron_dose_mg, data = metadata_df2, by = "terms")
bray_station_adonis_terms2

## 2. Run with by = "margin" for marginal p-values
bray_station_adonis_margin <- adonis2(scaled_bray_dist ~ Iron_dose_mg, data = metadata_df2, by = "margin")
bray_station_adonis_margin
```

```{r Principal Coordinates Analysis}
### SORENSEN 
# First, calculate PCoA with Soresen
scaled_soren_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE)

# Take a quick look
str(scaled_soren_pcoa)

# Plot it: Sorensen PCoA  
sample_data(scaled_physeq)$Iron_intake <- sample_data(scaled_physeq)$`Iron intake`

sorensen_pcoa_plot <- plot_ordination(
  physeq = scaled_physeq,
  ordination = scaled_soren_pcoa,
  color = "Iron_intake",
  shape = "Group",
  title = "Sorensen PCoA"
) +
  geom_point(size = 5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right")
sorensen_pcoa_plot
```
```{r Bray-Curtis PCoA }
### Bray-Curtis 
# Second, calculate PCoA with Bray-Curtis
scaled_bray_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = FALSE)

# Plot it: Bray-Curtis PCoA 
bray_pcoa_plot <- 
  plot_ordination(
    physeq = scaled_physeq,
    ordination = scaled_bray_pcoa,
  color = "Iron_intake",
  shape = "Group",
  title = "Bray-Curtis PCoA"
) +
  geom_point(size = 5, alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right")

# Show the plots 
sorensen_pcoa_plot + bray_pcoa_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

```{r stats}
# Bray-Curtis Betadisper - Station 
dispr_bray_station <- betadisper(scaled_bray_dist, metadata_df2$Iron.intake)
permutest(dispr_bray_station)

```

