---
title: "Beta-Diversity"
output: html_document
author: "Karina Zhang"
date: "2026-04-15"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inputs:
- preprocessed_physeq: asv, count table, metadata, taxonomy

# Outputs
- visualize data in ordinations (PCoA or NMDs)
-Calculate stats (PERMANOVA and betadisper)

```{r Load data and packages}
pacman::p_load(tidyverse, phyloseq, patchwork, vegan,
               install = FALSE)

#load data
load("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/Data/03_Preprocessing/preprocessed_phyloseq.RData")
preprocessed_physeq

source("/local/workdir/yz3272/Biodiversity_Salinity_16S/code/functions.R")

# Intuition check on seq depth
min(sample_sums(preprocessed_physeq))
```

###INTERPRETATION 1: 
My scientific question is whether feeding mice a low-iron diet versus a high-iron diet is associated with differences in the gut microbial community profile. My hypothesis is that greater dietary iron intake results in more unabsorbed iron reaching the murine colon, which may selectively promote or suppress the growth of specific microbial taxa. Consequently, the microbial community structure is expected to differ between diet groups. Because this hypothesis concerns changes in the abundance of taxa rather than simply their presence or absence, this analysis primarily relates to abundance-weighted metrics (relative abundance).

```{r Create metadata_df}
metadata_df <- sample_data(preprocessed_physeq) %>%
  data.frame() %>%
  filter(Group == "HC", Iron.intake %in% c("50", "500"))

physeq_HC <- prune_samples(rownames(metadata_df), preprocessed_physeq)
physeq_HC <- prune_taxa(taxa_sums(physeq_HC) > 0, physeq_HC)
```

```{r Explore the Raw Read Counts}
# Drop taxa with zero counts after subsetting
physeq_HC <- prune_taxa(taxa_sums(physeq_HC) > 0, physeq_HC)

# calculate read counts per sample 
raw_TotalSeqs_df <- 
  physeq_HC %>%
  sample_sums() %>%
  data.frame()

# Rename the column 
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs"

# Add sample IDs as a column
raw_TotalSeqs_df$SampleID <- rownames(raw_TotalSeqs_df)

# Calculate number of observed ASVs per sample
otu_mat <- as(otu_table(physeq_HC), "matrix")

# Make sure samples are rows
if (taxa_are_rows(physeq_HC)) {
  otu_mat <- t(otu_mat)
}

num_ASVs_per_sample <- rowSums(otu_mat > 1)

# add a new column of num_ASVs (RAW, non-normalized # ASVs per sample)
raw_TotalSeqsASVs_df <- 
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = num_ASVs_per_sample)

# Plot histogram of seq depth 
rawSeq_histogram <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 30) + 
  scale_x_continuous(limits = c(0, 70000)) + 
  labs(title = "Raw Seq Depth Histogram: HC50 and HC500") + 
  theme_bw()

# Plot Seq Depth versus num_ASVs
rawSeq_vs_numASV_plot <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 70000)) +
  geom_smooth(method = "lm") + 
  labs(title = "Seq Depth vs # ASVs: HC50 and HC500") + 
  theme_bw()

# is it significant? 
summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))

# Put the two plots together
seq_depth <- rawSeq_histogram + rawSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")

ggsave("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/Figures/05_Ordination/Seq_depth.png",
       plot = seq_depth, width = 12, height = 8, dpi = 600)
```
###INTERPRETATION 2:
The raw sequencing depths for HC_50 and HC_500 samples ranged from 21,000 to 58,000 reads, so read counts were scaled to the minimum sample depth to retain all samples while standardizing library size across groups. This scaling helps to reduce bias from unequal sequencing depth in downstream beta-diversity analyses, however, this way it discards a great portion of reads from deeper samples. Consequently, it may decrease the sensitivity to minor community differences and rare taxa abundance. 


```{r  Scale Read Counts}
min(sample_sums(preprocessed_physeq))

# Scale the reads 
scaled_physeq <- 
  physeq_HC %>%
  scale_reads(round = "matround")

# Look at it 
scaled_physeq

# Look at it more!
#View(data.frame(otu_table(preprocessed_physeq)))
#View(data.frame(otu_table(scaled_physeq)))

# Confirm seq read depth of scaled_physeq 
scaled_TotalSeqs_df <- 
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
scaled_TotalSeqsASVs_df <- 
  scaled_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(scaled_physeq) > 1))

# Plot it! 
scaledSeq_histogram <- 
  scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  labs(title = "Scaled Seq Depth Histogram") + 
  scale_x_continuous(limits = c(0, 70000)) + 
  theme_bw()
scaledSeq_histogram

# Scaling always gives us a range! 
# MInimum sequences
min(scaled_TotalSeqs_df$TotalSeqs) 

# Total Range 
range(scaled_TotalSeqs_df) # a range happen due to the scaling factor differ for each sample
# How many is that range? 
range_seqs <- (max(scaled_TotalSeqs_df$TotalSeqs) - min(scaled_TotalSeqs_df$TotalSeqs))
range_seqs
range_seqs/max(scaled_TotalSeqs_df$TotalSeqs)

#View(scaled_TotalSeqs_df)

# Set the axis min and max for the next plot 
max_y <- max(scaled_TotalSeqs_df$TotalSeqs) + 10
min_y <- min(scaled_TotalSeqs_df$TotalSeqs) - 10

# Now, draw the plot 
scaledSeq_vs_numASV_plot <- 
  scaled_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(min_y, max_y)) +
  #geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(title = "Scaled: Seq Depth vs # ASVs")

# Put the two plots together.
scaled_seq <- scaledSeq_histogram + scaledSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")

ggsave("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/Figures/05_Ordination/scaled_seq.png",
       plot = scaled_seq, width = 12, height = 8, dpi = 600)
```

###INTERPRETATION 3: 
After scaling, the total number of sequences across all HC samples was brought into the same general range, between 21,440 and 21,474 reads (a difference of only 34 reads, ~0.16%). The association between number of ASVs and read counts was flatter, therefore, downstream beta diversity analysis will like reflect true community differences instead of sequencing depth related artifacts.


```{r}
save(scaled_physeq, file = "/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/Data/05_Ordination/scaled_physeq.RData")
```


```{r Calculating dissimilarity}

sample_data(scaled_physeq)$Iron.intake <- factor(
  sample_data(scaled_physeq)$Iron.intake,
  levels = c(50, 500),
  labels = c("HC50", "HC500")
)

# Bray-Curtis Dissimiliarty
scaled_bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)
str(scaled_bray_dist)

bray_permanova <- adonis2(
  scaled_bray_dist ~ Iron.intake,
  data = metadata_df)
bray_permanova

### for exploration, run Sorensen to test the persence of major memebers:
# Sorensen Dissimiliarty
scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

sorensen_permanova <- adonis2(
  scaled_sorensen_dist ~ Iron.intake,
  data = metadata_df
)
sorensen_permanova

# What does it look like? 
class(scaled_sorensen_dist)

str(scaled_sorensen_dist)
```

### INTERPRETATION 4:
Given the fact that my question asks whether HC500 and HC50 differ in gut microbial community composition in terms of taxon abundances, not just presence or absence, Bray–Curtis would be the best fit as we do not have phylogenetic tree. Bray–Curtis is the most appropriate one because it is abundance-weighted and therefore better capture the dissimilartiy in the asked question. Sorensen Dissimiliarty was also explored, and it had a p-value of 0.342, again, suggesting the 10-fold diff in non-heme iron diet did not lead to sig. diff in the microbial taxa diversity between 2 groups.


```{r Principal Coordinates Analysis}
scaled_bray_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = FALSE)

# Plot it: Bray-Curtis PCoA 
bray_pcoa_plot <- plot_ordination(physeq = scaled_physeq,
                                  ordination = scaled_bray_pcoa) +
  geom_point(aes(color = factor(Iron.intake)),
             size = 4.5, alpha = 0.85) +
  stat_ellipse(aes(color = factor(Iron.intake)),
               type = "t", linewidth = 0.9) +
  scale_color_manual(values = c("HC50" = "#2C7FB8", "HC500" = "#D95F0E"),
                     labels = c("50 mg/kg", "500 mg/kg"),
                     name = "Iron intake") +
  labs(title = "Bray-Curtis PCoA",
       subtitle = "PERMANOVA p = 0.071") +
  theme_classic(base_size = 16)

ggsave("/local/workdir/yz3272/BIOMI6300_IronMicrobiotaProject/Figures/05_Ordination/Bray_curtis_plot.png",
       plot = bray_pcoa_plot, width = 12, height = 8, dpi = 600)

# Calculate PCoA with Sorensen dissimilarity
scaled_sorensen_pcoa <- 
  ordinate(
    physeq = scaled_physeq,
    method = "PCoA",
    distance = "bray",
    binary = TRUE
  )

# Plot it: Sorensen PCoA 
sorensen_pcoa_plot <- 
  plot_ordination(
    physeq = scaled_physeq,
    ordination = scaled_sorensen_pcoa,
    color = "Iron.intake",
    title = "Sorensen PCoA"
  ) + 
  geom_point(size = 5, alpha = 0.5, aes(color = Iron.intake)) + 
  theme_bw() + 
  theme(legend.position = "right")

# Show the plots 
sorensen_pcoa_plot+ bray_pcoa_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

###INTERPRETATION 5: 
The Bray–Curtis PCoA shows that the first two axes explain 33.4% and 14.7% of variation in the microbial community composition. HC_50 and HC_500 groups overlap substantially with no clustering observed, suggesting different iron dose (10-fold diff in non-heme content) has little effect on the gut microbiome diversity; this is also confirmed with PERMANOVA having a p-value of 0.1. The Sorensen plot did not capture as much diff as bray crutis PCoA plot, but basically telling the same story that the beta diverstiy did not diff sig. between groups.

```{r NMD plot}
# Up till here, i don't think NMD plot is necessary given my PCoA explains~48% of the distribution diff. just plotting to learn the code
scaled_bray_nmds <- 
  ordinate(
    physeq = scaled_physeq,
    method = "NMDS",
    distance = "bray",
    binary = "FALSE")

bray_nmds_plot <- 
  plot_ordination(
    physeq = scaled_physeq,
    ordination = scaled_bray_nmds,
    color = "Iron.intake",
    title = "Bray-Curtis NMDS"
  ) +
  geom_point(size = 5, alpha = 0.7) +
  theme_bw() +
  theme(legend.position = "right")

# Calculate NMDS with Sorensen dissimilarity
scaled_sorensen_nmds <- 
  ordinate(
    physeq = scaled_physeq,
    method = "NMDS",
    distance = "bray",
    binary = TRUE
  )

# Plot it: Sorensen NMDS
sorensen_nmds_plot <- 
  plot_ordination(
    physeq = scaled_physeq,
    ordination = scaled_sorensen_nmds,
    color = "Iron.intake",
    title = "Sorensen NMDS"
  ) +
  geom_point(size = 5, alpha = 0.7) +
  theme_bw() +
  theme(legend.position = "right")


# Show the plots 
sorensen_nmds_plot + bray_nmds_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

###INTERPRETATION 6:
Both NMDS plots show no seperations between HC_50 and HC_500 samples, indicating no sig. differences in gut microbial community composition by iron intake. The Sorensen NMDS suggests broadly similar microbial members between groups. The Bray–Curtis NMDS shows a little bit seperation, implying such treatment effect is more likely causing minor shifts in relative abundance than major taxonomic turnover.However, the stress value for NMDs is greater than 0.2 suggest the limitation of this distribution method, PCoA would be better representing the results.


```{r 4 plots together}
# Show plot
sorensen_pcoa_plot + bray_pcoa_plot + 
sorensen_nmds_plot + bray_nmds_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

###INTERPRETATION 7: Plot both the PCoA and NMDS together. Which of the two ordination methods best represents your data and why? Consider the strengths and limitations of each method. For example, the variance explained by each PCoA axis versus the stress value of the NMDS could be used to justify your choice. Do both methods support the same biological conclusions?

Among these 4 plots, PCoA best represents this data set because the Bray–Curtis PCoA explains a relatively high proportion of variation in the first two axes (48.1%), making the dominant beta-diversity patterns easier to interpret, whereas NMDS relies on stress values rather than variance explained. I think Bray–Curtis PCoA is preferable as the primary ordination, with NMDS serving as a secondary secreening for hidden patterns.


###INTERPRETATION 8: 
The PERMANOVA based on Bray–Curtis distancing suggested iron intake explained only ~3.5% of the variation in gut microbial community composition between HC_50 and HC_500 samples (R² = 0.035). This was not statistically significant (F = 1.75, p = 0.092), although the marginal p-value may suggest a weak trend toward differences between groups. More to be explored in differential abundance of taxa.Based on this beta diveristy, i can accpet my null hypo that the microbial diversity is the same in 50 and 500ppm iron groups.

```{r stats}
# Bray-Curtis Betadisper - Station 
dispr_bray_station <- betadisper(scaled_bray_dist, metadata_df$Iron.intake)
permutest(dispr_bray_station)
```
###INTERPRETATION 9: 
A permutation test based on Bray–Curtis distances was non-significant (F = 0.92, p = 0.342), indicating similar within-group dispersion between HC_50 and HC_500. Therefore, the previously observed PERMANOVA differences are not driven by unequal group variability (although not sig.).

###INTERPRETATION 10: 
Overall, results seen here suggested that gut microbial community composition did not differ significantly between HC_50 and HC_500 mice, and within group variability was similar. Therefore, a 10-fold difference in dietary non-heme iron did not produce a clear overall shift in beta-diversity, though subtle taxon-level changes may still be present.



